Expression QTL hotspot dissection with mediation analysis and pleiotropy testing

### 2019-08-12 analysis/paper/Rmd/methods.Rmd
added details of annotation info for heatmaps

@keller2018genetic performed univariate QTL mapping for all expression traits. They
defined an expression trait hotspot to be a 4-Mb region that affects at least 100 nonlocal 
expression traits.
They identified five expression trait hotspots in their
study of pancreatic islet gene expression in 378 Diversity Outbred mice .

@@

We created scatter plots with the `ggplot2` package [@ggplot2] for the R
statistical computing environment [@r]. We plotted all local-nonlocal pairs for each hotspot. We highlighted in blue those points that correspond to pairs involving the putative mediator that @keller2018genetic identified.

We created heatmaps to examine patterns among pleiotropy test results in each hotspot. Our analysis used both 
local and nonlocal traits.
We represented the pleiotropy test statistics as a symmetric matrix, where each row was one expression
trait and each column was an expression trait. 
Excepting the cells on the diagonal, each cell, then, contained the pleiotropy test statistic
from the analysis involving the row's expression trait and the column's expression trait.
To create a heatmap, we first performed two hierarchical clusterings of rows. 
One hierarchical clustering involved only the rows and columns containing local traits, while the other contained only 
rows and columns containing nonlocal traits.
We calculated the Euclidean distances between all row pairs (in the respective 
submatrices) before clustering. Hierarchical clustering was performed using the complete 
linkage method. Initially, each row is assigned to its own cluster. An iterative 
algorithm then combines the two most similar clusters at each step. Distances between 
clusters were computed with the Lance-Williams dissimilarity update formula for complete 
linkage [@lance1967general]. Lastly, we arranged the rows according to optimal leaf ordering [@hahsler2008getting;@dendextend]. 
We then combined the
local and nonlocal traits into a single heatmap for each hotspot. 
To achieve coloring contrasts in the heatmaps, we
set the color spectrum limits at 0 (dark blue) and 5 (yellow). Pleiotropy test statistics above 5, too, are 
represented by
yellow. We annotated rows by a collection of columns on the right-hand side of the heatmap. This binary matrix 
consists of columns that summarize mediation results. Each column represents a local trait. If a row 
corresponds to a nonlocal trait that is mediated by the local trait (*i.e.*, with mediation LOD 
difference greater than 1.5), then the cell is colored dark green; 
otherwise, the cell is light green. We didn't examine the possibility that a local trait might 
mediate the relationship between a QTL and another local trait; thus, all local trait rows are 
colored light green. The last annotation column and the annotation row above the heatmap both 
designate local traits with dark blue and nonlocal traits with light blue.

>>>>

@keller2018genetic performed univariate QTL mapping for all expression traits. They
defined an expression trait hotspot to be a 4-Mb region that affects at least 100 nonlocal 
expression traits.
They identified five expression trait hotspots in their
study of pancreatic islet gene expression in 378 Diversity Outbred mice.

@@

We created scatter plots with the `ggplot2` package [@ggplot2] for the R
statistical computing environment [@r]. We plotted pleiotropy test statistics against 
mediation LOD differences for all local-nonlocal pairs at each hotspot. We highlighted in blue 
those points that correspond to pairs involving the putative mediator that @keller2018genetic 
identified. We used the same results in plots of pleiotropy test statistics against mediation LOD difference for analyses involving a single local trait.

We created heatmaps to examine patterns among pleiotropy test results in each hotspot. Our analysis used both 
local and nonlocal traits.
We represented the pleiotropy test statistics as a symmetric matrix, where each row was one expression
trait and each column was an expression trait. 
Excepting the cells on the diagonal, each cell, then, contained the pleiotropy test statistic
from the analysis involving the row's expression trait and the column's expression trait.
To create a heatmap, we first performed two hierarchical clusterings of rows. 
One hierarchical clustering involved only the rows containing local traits, while the other contained only 
rows corresponding to nonlocal traits.
We calculated the Euclidean distances between all row pairs (in the respective 
submatrices) before clustering. Hierarchical clustering was performed using the iterative complete 
linkage method. Initially, each row is assigned to its own cluster. The iterative 
algorithm then combines the two most similar clusters at each step. Distances between 
clusters were computed with the Lance-Williams dissimilarity update formula for complete 
linkage [@lance1967general]. Lastly, we arranged the rows according to optimal leaf ordering [@hahsler2008getting;@dendextend]. 
We then combined the
local and nonlocal traits into a single heatmap for each hotspot. 
To achieve coloring contrasts in the heatmaps, we
set the color spectrum limits at 0 and 5. 
We annotated rows with six pieces of information: 1. mediation LOD differences, 2. mediation 
LOD difference proportions, 3. univariate QTL peak positions, 4. estimated allele effects 
vector at the univariate peak position, 5. univariate LOD score, and 6. whether the row 
corresponds to a local trait. For clarity of presentation, we also indicated those columns that  
correspond to a local traits. We didn't examine the possibility that a local trait might 
mediate the relationship between a QTL and another local trait.
